---
title: "Untitled"
editor: source
---

# Visualising time series data with `timetk`

## Packages

```{r, warning=FALSE, message=FALSE}
library("timetk")
library("tidyquant")
library(tidyverse)
library(denguedatahub)
library(anomalize)
library(tsibble)
library(fable)
library(fabletools) 
library(feasts) #STL decomposition
```

## Data

```{r}
data(srilanka_weekly_data)
colombo <- srilanka_weekly_data |>
  filter(district == "Colombo")
# create a proper date index column
colombo <- srilanka_weekly_data |>
  filter(district == "Colombo") |>
  mutate(
    start.date = mdy(start.date)
  )
```

## Time Series Plot

```{r}

colombo |>
  plot_time_series(
    .date_var = start.date,
    .value = cases,
    .interactive = FALSE
  ) +
  labs(
    title = "Weekly Dengue Cases in Colombo",
    x = "Year",
    y = "Cases"
  )
```

```{r}
data(srilanka_weekly_data)
colombo <- srilanka_weekly_data |>
  filter(district == "Colombo")
# create a proper date index column
colombo <- srilanka_weekly_data |>
  filter(district == "Colombo") |>
  mutate(
    start.date = mdy(start.date)
  )
```

## Seasonal plot

```{r, fig.height=15, fig.width=10}
colombo |>
  plot_seasonal_diagnostics(
    .date_var = start.date,
    .value = cases,
    .interactive = FALSE
  )
```

## STL decomposition


What do you think about the plot? Is this correct?

```{r, fig.height=15, fig.width=10}
colombo |>
  plot_stl_diagnostics(
    .date_var = start.date,
    .value = cases,
    .interactive = FALSE
  )
```


Let's fix the issue using tsibble and fabletools

What is wrong with the following code?

```{r}
colombo_tsibble <- srilanka_weekly_data |>
  filter(district == "Colombo") |>
  mutate(
    start.date = mdy(start.date)
  ) |>
  arrange(start.date) |>
  as_tsibble(index = start.date)
colombo_tsibble
```

Let's fix the issue

```{r}
colombo_tsibble <- srilanka_weekly_data |>
  filter(district == "Colombo") |>
  mutate(
    start.date = mdy(start.date),
    year_week = yearweek(start.date)
  ) |>
  arrange(year_week) |>
  as_tsibble(index = year_week)
colombo_tsibble
interval(colombo_tsibble)
```

```{r}
colombo_stl <- colombo_tsibble |>
  model(
    STL(cases ~ season(period = 52))
  )
colombo_stl
```

```{r}
colombo_components <- colombo_stl |>
  components()
colombo_components
```

```{r}
colombo_components |>
  autoplot()
```

Let's fix the issue with timetk

```{r, fig.height=15, fig.width=10}
colombo %>%
  plot_stl_diagnostics(
    .date_var = start.date,
    .value = cases,
    .frequency = 52,
    .trend = 104,
    .interactive = FALSE
  )
```

## Anomaly Detection

```{r, fig.height=15, fig.width=10}
colombo %>%
  time_decompose(
    cases,
    method = "stl",
    frequency = 52
  ) %>%
  anomalize(
    remainder,
    method = "iqr"
  ) %>%
  time_recompose() %>%
  plot_anomalies(time_recomposed = TRUE)

```

## Calendar Heatmap

```{r}
colombo |>
  plot_time_series(
    .date_var = start.date,
    .value = cases,
    .color_var = year(start.date),
    .interactive = TRUE
  )
```

or

```{r}
library(ggplot2)

colombo |>
  mutate(
    Year = year(start.date),
    Week = isoweek(start.date)
  ) |>
  ggplot(aes(Week, factor(Year), fill = cases)) +
  geom_tile() +
  labs(
    title = "Weekly Dengue Cases Heatmap",
    x = "Week",
    y = "Year"
  )
```

## Rolling Mean

```{r}
library(dplyr)
library(slider)
library(timetk)

colombo |>
  mutate(
    MA52 = slide_dbl(cases,
                     mean,
                     .before = 51,
                     .complete = TRUE)
  ) |>
  plot_time_series(
    .date_var = start.date,
    .value = cases,
    .interactive = FALSE
  ) +
  geom_line(aes(y = MA52), color = "blue")
```

## Time Series Signature Exploration

```{r}
colombo |>
  tk_augment_timeseries_signature(start.date) |>
  glimpse()
```

## Additional Reading

[See the plot](https://business-science.github.io/timetk/articles/TK04_Plotting_Time_Series.html)
