---
title: "Movement Data Visualisation"
editor: source
toc: true
---

# Simulate Movement Data

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
set.seed(123)
n <- 120
movement <- data.frame(
  Time = seq.POSIXt(
    from = as.POSIXct("2025-01-01 08:00"),
    by = "30 min",
    length.out = n
  )
)

movement$X <- cumsum(rnorm(n, mean = 2, sd = 4))
movement$Y <- cumsum(rnorm(n, mean = 1, sd = 3))

head(movement)
```

# Static Plots

## Without any background

```{r}
library(ggplot2)

ggplot(movement, aes(x = X, y = Y)) +
  geom_path(colour = "steelblue", linewidth = 0.7, alpha = 0.8) +
  geom_point(colour = "red", size = 2, alpha = 0.7) +
  labs(
    title = "Simulated Animal Movement",
    x = "X Coordinate",
    y = "Y Coordinate"
  ) +
  theme_minimal()
```

## With map background

```{r}
# Install packages if needed
# install.packages(c("sf", "maptiles", "ggplot2", "dplyr", "ggspatial"))

library(dplyr)
library(ggplot2)
library(sf)
library(maptiles)
library(ggspatial)

#-------------------------------------------------
# Simulate animal movement
#-------------------------------------------------

set.seed(123)

n <- 120

movement <- data.frame(
  Time = seq.POSIXt(
    from = as.POSIXct("2025-01-01 08:00"),
    by = "30 min",
    length.out = n
  )
)

# Simulated random walk
movement$dx <- rnorm(n, mean = 0.001, sd = 0.003)
movement$dy <- rnorm(n, mean = 0.0005, sd = 0.002)

# Starting location (Colombo)
start_lon <- 79.8612
start_lat <- 6.9271

movement$Longitude <- start_lon + cumsum(movement$dx)
movement$Latitude  <- start_lat + cumsum(movement$dy)

#-------------------------------------------------
# Convert to sf object
#-------------------------------------------------

pts <- st_as_sf(
  movement,
  coords = c("Longitude", "Latitude"),
  crs = 4326,
  remove = FALSE
)

#-------------------------------------------------
# Download OpenStreetMap tiles
#-------------------------------------------------
# downloads map tiles from OpenStreetMap for the area covered by your movement points.
osm <- get_tiles(
  pts,
  provider = "OpenStreetMap",
  zoom = 13
)

#-------------------------------------------------
# Plot
#-------------------------------------------------

ggplot() +
  layer_spatial(osm) +
  geom_path(
    data = movement,
    aes(Longitude, Latitude),
    colour = "blue",
    linewidth = 1
  ) +
  geom_point(
    data = movement,
    aes(Longitude, Latitude),
    colour = "red",
    size = 2
  ) +
  labs(
    title = "Simulated Animal Movement",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal()
```

# Interactive Visualisation

```{r}
# Install if needed
# install.packages(c("sp", "mapview"))

library(sp)
library(mapview)
library(dplyr)

#---------------------------------------
# Simulated movement data
#---------------------------------------

set.seed(123)

n <- 120

movement <- data.frame(
  Time = seq.POSIXt(
    from = as.POSIXct("2025-01-01 08:00"),
    by = "30 min",
    length.out = n
  )
)

# Simulated longitude and latitude
movement$Longitude <- 79.8612 + cumsum(rnorm(n, 0.001, 0.003))
movement$Latitude  <- 6.9271 + cumsum(rnorm(n, 0.0005, 0.002))


# Create an animal ID
movement$Animal <- "Caribou_1"


#---------------------------------------
# Convert data frame to spatial points
#---------------------------------------

movement.sp <- movement

coordinates(movement.sp) <- ~ Longitude + Latitude


# Assign coordinate reference system
proj4string(movement.sp) <- CRS("+proj=longlat +datum=WGS84")


#---------------------------------------
# Interactive map
#---------------------------------------

mapview(
  movement.sp,
  zcol = "Animal",
  legend = TRUE,
  cex = 3,
  lwd = 1,
  map.types = "Esri.WorldImagery"
)
```

```{r}
mapview(
  movement.sp,
  zcol = "Animal",
  legend = TRUE,
  cex = 3,
  map.types = "Esri.WorldImagery"
) +
mapview(
  SpatialLines(
    list(
      Lines(
        list(
          Line(coordinates(movement.sp))
        ),
        ID = "Caribou_1"
      )
    ),
    proj4string = CRS("+proj=longlat +datum=WGS84")
  ),
  color = "red"
)
```


This code creates a spatial line (trajectory) from your simulated movement points and displays it interactively using mapview.


                    *-------------*


### 1. Extract coordinates
coordinates(movement.sp)

movement.sp is a spatial points object.

It contains:

Longitude  Latitude

79.8612     6.9271

79.8625     6.9280

79.8640     6.9295
...

coordinates() extracts only the coordinate matrix:

        Longitude Latitude
        
[1,]     79.8612   6.9271

[2,]     79.8625   6.9280

[3,]     79.8640   6.9295


### 2. Create a line from coordinates

Line(coordinates(movement.sp))

Line() is an sp package function.

It takes a sequence of coordinates and joins them in order.

Example:

Input:

(79.861, 6.927)
(79.862, 6.928)
(79.864, 6.930)

Output:

(79.861,6.927) -------- (79.862,6.928) -------- (79.864,6.930)

This represents the animal's movement trajectory.

### 3. Convert Line into a Lines object

Lines(
  list(
    Line(coordinates(movement.sp))
  ),
  ID = "Caribou_1"
)
Why another function?

Line() creates one line.

Lines() creates a spatial line object with an identifier.

The structure is:

Example:

Caribou_1 → Line 1
Caribou_2 → Line 2
Caribou_3 → Line 3

The ID allows multiple animals to have different paths.

### 4. Create SpatialLines object

```r
SpatialLines(
  list(
    Lines(...)
  ),
  proj4string = CRS("+proj=longlat +datum=WGS84")
)
```

A Lines object only contains geometry.

SpatialLines() adds:

Coordinates

Coordinate reference system (CRS)

The CRS:

CRS("+proj=longlat +datum=WGS84")

means:

coordinates are longitude/latitude
based on the WGS84 Earth reference system
the same system used by GPS devices

So R knows:

79.8612 = longitude
6.9271  = latitude

### 5. Display using mapview

mapview(
  SpatialLines(...),
  color = "red"
)

mapview() creates an interactive Leaflet map.

The output:

satellite background
red movement trajectory
zoom and pan controls
Complete workflow
movement data frame

Longitude Latitude
    |        |
    ↓        ↓

SpatialPoints
    |
    ↓
coordinates()

    |
    ↓

Line()
(connect points)

    |
    ↓

Lines()
(add ID)

    |
    ↓

SpatialLines()
(add CRS)

    |
    ↓

mapview()
(interactive map)

In animal movement analysis, this line represents the animal's track or trajectory, while the original points represent individual GPS observations.

# Other map types

## Example 1: OpenStreetMap (road map)

```{r}
mapview(
  movement.sp,
  zcol = "Animal",
  map.types = "OpenStreetMap"
)
```

## Example 2: Satellite imagery

```{r}
mapview(
  movement.sp,
  zcol = "Animal",
  map.types = "Esri.WorldImagery"
)
```

## Example 3: Topographic map

```{r}
mapview(
  movement.sp,
  zcol = "Animal",
  map.types = "Esri.WorldTopoMap"
)
```


## Example 4: Dark background

```{r}
mapview(
  movement.sp,
  zcol = "Animal",
  map.types = "CartoDB.DarkMatter"
)
```

## Example 5: Multiple basemaps (switch interactively)

You can provide several maps:

```{r}
mapview(
  movement.sp,
  zcol = "Animal",
  map.types = c(
    "OpenStreetMap",
    "Esri.WorldImagery",
    "Esri.WorldTopoMap"
  )
)
```


Then the map viewer will allow you to switch between:

Road map
Satellite map
Terrain map

For animal movement visualization, I would usually use:

map.types = "Esri.WorldImagery"

for ecological interpretation (forest, rivers, habitat), and

map.types = "OpenStreetMap"

for teaching movement trajectories because roads and landmarks are easier to identify.